---
title: "Google Data Analytics Capstone"
author: "Cezar Trevino"
date: "8/19/2021"
output: html_document
---

# Introduction
This is a detailed analysis of case study 1 "How does a bike-share navigate speedy success?" within the Google Data Analytics certificate. This analysis involves data from Cyclistic's previous 12 months (July 2020 - June 2021). The data has been made available by Motivate International Inc. under this [license](https://www.divvybikes.com/data-license-agreement).

<br>

### Scenario
Cyclistic is a bike-share program and the director of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore, your team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, your team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve your recommendations, so they must be backed up with compelling data insights and professional data visualizations.

<br>

# Objective
The objective of this case study I am aiming to tackle is analyze how casual members (casuals) and annual members (annuals) differ. Figuring out how casuals and annuals differ will aid in the goal of maximizing annual memberships.

<br>

# Sources
12 datasets were made available from previous months of Cyclistic's bike ride data.The datasets contain data from every logged bike ride that occured during the used months. The datasets combined contained around 4 million observations

<br>

# Cleaning and Visualizing
I started to clean in excel but the vast amount of data was making the process of cleaning within excel unrealistic. I choose to switch to R to better learn and get comfortable with the language.

<br>

Load packages
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)
library(ggthemes)
```
<br>

Load datasets
```{r}
july2020 <- read.csv("C:/Users/bigda/Documents/datasetsCapstone/202007-divvy-tripdata.csv")
august2020 <- read.csv("C:/Users/bigda/Documents/datasetsCapstone/202008-divvy-tripdata.csv")
september2020 <- read.csv("C:/Users/bigda/Documents/datasetsCapstone/202009-divvy-tripdata.csv")
october2020 <- read.csv("C:/Users/bigda/Documents/datasetsCapstone/202010-divvy-tripdata.csv")
november2020 <- read.csv("C:/Users/bigda/Documents/datasetsCapstone/202011-divvy-tripdata.csv")
december2020 <- read.csv("C:/Users/bigda/Documents/datasetsCapstone/202012-divvy-tripdata.csv")
january2021 <- read.csv("C:/Users/bigda/Documents/datasetsCapstone/202101-divvy-tripdata.csv")
february2021 <- read.csv("C:/Users/bigda/Documents/datasetsCapstone/202102-divvy-tripdata.csv")
march2021 <- read.csv("C:/Users/bigda/Documents/datasetsCapstone/202103-divvy-tripdata.csv")
april2021 <- read.csv("C:/Users/bigda/Documents/datasetsCapstone/202104-divvy-tripdata.csv")
may2021 <- read.csv("C:/Users/bigda/Documents/datasetsCapstone/202105-divvy-tripdata.csv")
june2021 <- read.csv("C:/Users/bigda/Documents/datasetsCapstone/202106-divvy-tripdata.csv")

```

<br>

Making all dataframes have the same column names and datatype. Convert id columns from int to chr
```{r}
#july2020
july2020$start_station_id <- as.character(july2020$start_station_id)
july2020$end_station_id <- as.character(july2020$end_station_id)

#august2020
august2020$start_station_id <- as.character(august2020$start_station_id)
august2020$end_station_id <- as.character(august2020$end_station_id)

#september2020
september2020$start_station_id <- as.character(september2020$start_station_id)
september2020$end_station_id <- as.character(september2020$end_station_id)

#october2020
october2020$start_station_id <- as.character(october2020$start_station_id)
october2020$end_station_id <- as.character(october2020$end_station_id)

#november2020
november2020$start_station_id <- as.character(november2020$start_station_id)
november2020$end_station_id <- as.character(november2020$end_station_id)
```

<br>

Merging all dataframes into one single dataframe
```{r}
full_dataset <- bind_rows(july2020, august2020, september2020, october2020, november2020,
                          december2020, january2021, february2021, march2021, april2021,
                          may2021, june2021)
```

<br>

Removing all 480 rows with a zero duration
```{r}
full_dataset <- full_dataset[!(full_dataset$ride_length == "0:00:00"),]
```

<br>

Removing day_of_week column I made in excel because it is resulting +3 million NA values
```{r}
full_dataset <- full_dataset %>% 
  select(rideable_type, started_at, ended_at, start_station_name, end_station_name, member_casual, ride_length)
```

<br>

Changing started_at and ended_at to datetime
```{r}
full_dataset$start_date <- mdy_hm(full_dataset$started_at)

full_dataset$end_date <- mdy_hm(full_dataset$ended_at)
```

<br>

Changing ride_length to time format
```{r}
full_dataset$ride_duration <- strptime(full_dataset$ride_length, "%H:%M:%S")
```

<br>

Removing unnecessary columns
```{r}
full_dataset <- full_dataset %>% 
  select(rideable_type, start_station_name, end_station_name, member_casual, start_date, end_date, ride_duration)
```

<br>

Creating a day of week column where the day of week corresponds to a number. Ex: 1 = Sunday, 2 = Monday ect.
```{r}
full_dataset$day_of_week <- wday(full_dataset$start_date)
```

<br>

Replacing blank cells with NA
```{r}
full_dataset$start_station_name[full_dataset$start_station_name == ""] <- NA

full_dataset$end_station_name[full_dataset$end_station_name == ""] <- NA

full_dataset$rideable_type[full_dataset$rideable_type == ""] <- NA

full_dataset$member_casual[full_dataset$member_casual == ""] <- NA
```

<br>

Viewing how many NA columns there are 
```{r}
colSums(is.na(full_dataset))
```

It shows the dataframe contains 282064 under start_station_name, 314904 under end_station_name and 13436 totaling 610,404 NA values. We will remove all NA values as it only accounts for around 14% of the 4,459,671 total dataframe and will lead to incorrect conclusions

<br>

Convert ride_duration from posixlt to posixct for na.omit
```{r}
full_dataset$ride_duration <- as.POSIXct(full_dataset$ride_duration)
```

<br>

Remove all NA columns 
```{r}
full_dataset <- na.omit(full_dataset)
```
Obs count went from 4,459,671 to 4,014,154

<br>

na.omit wowuld not take away all NA values in ride_duration so I used complete.cases to finish the job
```{r}
full_dataset$ride_duration <- complete.cases(full_dataset$ride_duration)
```

<br>

Creating a month column 
```{r}
full_dataset$month <- month(full_dataset$start_date)
```

<br>

Creating a year column 
```{r}
full_dataset$year <- year(full_dataset$start_date)
```

<br>

Removing rows where ride length are negative
```{r}
full_dataset <- full_dataset[!(full_dataset$start_station_name == "HQ QR" | full_dataset$ride_duration<0),]
```

<br>

Add a "ride_length" calculation to all_trips (in seconds)
```{r}
full_dataset$ride_length <- difftime(full_dataset$end_date,full_dataset$start_date)
```

<br>

Convert "ride_length" from Factor to numeric so we can run calculations on the data
```{r}
is.factor(full_dataset$ride_length)
full_dataset$ride_length <- as.numeric(as.character(full_dataset$ride_length))
is.numeric(full_dataset$ride_length)
```
<br>

Remove all ride_lengths with value of zero 
```{r}
full_dataset <- full_dataset[!(full_dataset$ride_length == 0),]
```

<br>

Make a column where ride_length is in minutes instead of seconds
```{r}
full_dataset$ride_length_minutes <- (full_dataset$ride_length/60)
```

<br>

View summary of ride_length in minutes
```{r}
summary((full_dataset$ride_length_minutes))
```
From the summary we we out of all rides logged the average ride duration is 23 minutes.

<br>

lets comapre the average duration or rides for each type of user 
```{r}
aggregate(full_dataset$ride_length_minutes ~ full_dataset$member_casual, FUN = mean)
```
So on average casual riders ride for longer than membership riders.

<br>

Analyze ridership data by type and weekday
```{r}
full_dataset %>% 
  mutate(weekday = wday(start_date, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length_minutes)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts
```
This data in useful but could be more useful in a visual format.

<br>

It might be useful to see what days of the week casuals ride and what days of the week members ride. Lets visualize this the data above to gain more insight to how each type or user rides throughout the week
```{r}
full_dataset %>% 
  mutate(weekday = wday(start_date, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + scale_y_continuous(labels = comma) +
  ggtitle("Number of rides throughout the week") +
  xlab("Day of the Week") +
  ylab("Number of Rides") +
  labs(fill="Rider Type") +
  theme_classic() + 
  scale_fill_manual(values = c("casual" = "#0063B2FF",
                               "member" = "#9CC3D5FF"))
```

From the graph above it shows causal memebrs tend to ride more promenantly during the weekends and members more during monday - friday. This brings about more questions, why? The way each rider type uses the service is a key factor to why they use the service the way they do.

<br>

In the summary we noticed that casuals on average ride longer than members, lets visualize ride times to get a clearer picture.
```{r}
full_dataset %>% 
  mutate(weekday = wday(start_date, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = (average_duration/60), fill = member_casual)) +
  geom_col(position = "dodge") +
  ggtitle("Average Ride Duration Throught The Week") +
  xlab("Day of the Week") +
  ylab("Minutes") +
  labs(fill="Rider Type") +
  theme_classic() +
  scale_fill_manual(values = c("casual" = "#0063B2FF",
                               "member" = "#9CC3D5FF"))
```

From the graph we see that there are in fact a difference in ride duration between the two types of riders. More analysis on diffrent type of data would need to be explored to fully understand this difference. We could draw inferences from the number of weekly riders throught the week graph and this one though. Looking at both graphs we see casual member tend to ride for longer and ride most frequently on the weekends. Could this be becauyse casual riders use the service for more leasure and members more for commuting? More qualitiave measure would need to be persued to find these answers such as surveys, ect.

<br>

Lets create a graph where it shows the releationship between each type of rider and what type of bike they choose to use the most.
```{r}
ggplot(full_dataset) + aes(x = full_dataset$rideable_type, fill = member_casual) + 
  geom_bar(position = "dodge") + scale_y_continuous(labels = comma) +
  ggtitle("Bike Type Usage") +
  xlab("Type of Bike") +
  ylab("Number of Bikes") +
  labs(fill="Rider Type") +
  theme_classic() +
  scale_fill_manual(values = c("casual" = "#0063B2FF",
                               "member" = "#9CC3D5FF"))
```

The graph above displays each riders types bike preference. We see that the most popular bike type is the docked type. Also, there seems to be a big discrepancy in classic bike where members tend to use classic bike more frequently than casual riders. These seems to be some motivation to look into why members prefer the classic bike much more than classic bikes. Electric bike are the least utilized, could that be because or price, or aesthetics? It could be useful to look into the reasons for the bike types usage as well.

<br>

Lets create a graph where it shows the number or rides between casuals and members each month.
```{r}
full_dataset$month_name <- month(full_dataset$start_date, label=TRUE)

full_dataset %>%  
  group_by(member_casual, month_name) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(member_casual, month_name)  %>% 
  ggplot(aes(x = month_name, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + 
  scale_y_continuous(labels = comma) +
  ggtitle("Bike Usage Each Month") +
  xlab("Month") +
  ylab("Number of Rides") +
  labs(fill="Rider Type") +
  theme_classic() +
  scale_fill_manual(values = c("casual" = "#0063B2FF",
                               "member" = "#9CC3D5FF"))
```


From the graph above we can see that the number of rides has a time where business is slower and business is more prominent. During the winters months such as December, January, and February are where the number or rides slow down drastically. If we look at the summer months we see the number of rides drastically increases. When we look at the winter months it seems there is a substantial gap between the number or members and causals who ue the service.

<br>

### Major Conclusions

1. Casual riders Tend to use the service on the weekends and Members more during the work week.
2. Business dips during the winter months and picks up during the spring and even more during the summer months. With that being said, during those winter months, members are the majority users of the service.
3. Docked bikes are most used, then comes classic bike, and finally, electirc bikes. Each bike type is used most by Members, but classic bike has the biggest discrpency. 
4. Electric bikes are under utilized relative to the other bike types.
5. When Casuals use the serive they, on average, ride for longer than members.

<br>

### What Could Be Done To Lure Casuals To Become Members?

1. Target casual riders for the objective of getting them to ride more during the week by providing promotions incentives. An example of this could be coupon codes for percentage discounts availiabe to be used during the weekdays that users can apply throguth the app.
2. Offer the first month free for new member during the time that casual riders are the most abundant which would be during the summer months.
3. If Cyclistic see an importance to imporve the uasge of electirc bike, Cyclistic could offer incentive to use that type of bike. These incentive could include use electric bike get a free ride up to a certian time limit.
4. To attract more Casuals to become Members, Cyclistic could add a memebership pass where it only applies to weekends, where Casuals ride most frequently. For instance, The pass would only be applicable to Saturday, and Sunday.